---
title: "Getting started with RTrino"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RTrino}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`RTrino` is a DBI backend for [Trino](https://trino.io), the distributed SQL
query engine. It talks to a Trino coordinator over the cluster's HTTP REST API,
so there is no driver to install and nothing to compile.

The code in this vignette is not evaluated: every example needs a reachable
cluster.

## Connecting

Trino's REST API is stateless, so a connection is really a bundle of settings:
the coordinator's address, who you are, and which catalog and schema unqualified
table names resolve against. `catalog` and `schema` are required.

```{r}
library(DBI)
library(RTrino)

con <- dbConnect(
  Trino(),
  host = "http://localhost",
  port = 8080,
  catalog = "hive",
  schema = "default"
)
```

`dbConnect()` calls `/v1/info` before returning, so a wrong address or a
rejected credential fails immediately rather than on your first query.

## Authentication

Each authentication method is a closure that attaches credentials to every
request. The credentials live inside that closure, never in the global
environment, and never in a slot you might print by accident.

Trino refuses to accept credentials over plain HTTP, so all three need an
`https://` host.

### Basic — LDAP or a password file

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.example.com",
  port = 443,
  user = Sys.getenv("TRINO_USER"),
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_basic(Sys.getenv("TRINO_USER"), Sys.getenv("TRINO_PASSWORD"))
)
```

### JWT — service accounts and pipelines

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.example.com",
  port = 443,
  user = Sys.getenv("TRINO_USER"),
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_jwt(Sys.getenv("TRINO_TOKEN"))
)
```

### OAuth2 client credentials — corporate SSO

`httr2` caches the token and renews it when it expires, so a long-running
session does not need to think about the token's lifetime.

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.example.com",
  port = 443,
  user = Sys.getenv("TRINO_USER"),
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_oauth2(
    client_id     = Sys.getenv("OAUTH_CLIENT_ID"),
    client_secret = Sys.getenv("OAUTH_CLIENT_SECRET"),
    token_url     = "https://auth.example.com/oauth/token"
  )
)
```

## Internal certificate authorities

A cluster whose certificate is signed by your own CA does not need certificate
checking turned off. Point `trino_ssl()` at the CA's PEM file and verification
keeps working:

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.internal",
  port = 443,
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_basic(Sys.getenv("TRINO_USER"), Sys.getenv("TRINO_PASSWORD")),
  ssl_options = trino_ssl(ca_bundle = "/etc/ssl/certs/internal-ca.pem")
)
```

`trino_ssl(verify = FALSE)` exists for a development cluster with a throwaway
certificate. It accepts *any* certificate, including one presented by an
attacker, so it warns on every request it is applied to.

## Querying

```{r}
sales <- dbGetQuery(con, "SELECT * FROM sales LIMIT 100")
```

For a result too large to hold in memory, take it in chunks. `dbFetch()`
follows Trino's pagination for you and buffers whatever a page delivered beyond
what you asked for:

```{r}
res <- dbSendQuery(con, "SELECT * FROM sales")
while (!dbHasCompleted(res)) {
  chunk <- dbFetch(res, n = 10000)
  # summarise, write out, accumulate...
}
dbClearResult(res)
```

Clearing a result that is still running sends a `DELETE` to the coordinator, so
abandoning a query also stops the cluster working on it.

## Big integers

Trino's `BIGINT` spans values an R double cannot hold exactly. By default
`RTrino` returns such columns as `bit64::integer64`, which is exact:

```{r}
dbGetQuery(con, "SELECT 9007199254740993 AS id")$id
#> integer64
#> [1] 9007199254740993
```

Choose a different representation at connection time with
`bigint = "numeric"` (a double, exact only up to 2^53) or
`bigint = "character"`.

## dplyr

With dbplyr 2.6.0 or newer installed, a connection carries a Trino SQL dialect
and `dplyr` verbs are translated to SQL Trino accepts:

```{r}
library(dplyr)

tbl(con, "sales") |>
  filter(year == 2026) |>
  group_by(region) |>
  summarise(total = sum(amount, na.rm = TRUE)) |>
  collect()
```

Trino-specific translations are applied where the ANSI defaults would not work:
casts name Trino types, `grepl()`/`gsub()` become `REGEXP_LIKE`/
`REGEXP_REPLACE`, and `median()`/`quantile()` become `APPROX_PERCENTILE`, which
is the percentile Trino offers over an arbitrary number of rows.

Inspect the generated SQL before running it with `dplyr::show_query()`, and the
cluster's plan with `dplyr::explain()`.

## Introspection

```{r}
dbListTables(con)
dbListFields(con, "sales")
dbExistsTable(con, "sales")
dbGetInfo(con)$db.version
```

## Session properties and other headers

Anything the protocol expects in a header can be added to every request through
`extra.headers` — Trino session properties, for instance:

```{r}
con <- dbConnect(
  Trino(),
  host = "http://localhost",
  port = 8080,
  catalog = "hive",
  schema = "default",
  extra.headers = list("X-Trino-Session" = "query_max_run_time=2h")
)
```

## Disconnecting

There is no server-side session to close, so `dbDisconnect()` only invalidates
the object; any later use of it is an error rather than a silent reconnect.

```{r}
dbDisconnect(con)
```
