---
title: "Getting started with RTrino"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RTrino}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`RTrino` is a DBI backend for [Trino](https://trino.io), the distributed SQL
query engine. It talks to a Trino coordinator over the cluster's HTTP REST API,
so there is no driver to install and nothing to compile.

The code in this vignette is not evaluated: every example needs a reachable
cluster.

## Connecting

Trino's REST API is stateless, so a connection is really a bundle of settings:
the coordinator's address, who you are, and which catalog and schema unqualified
table names resolve against. `catalog` and `schema` are required.

```{r}
library(DBI)
library(RTrino)

con <- dbConnect(
  Trino(),
  host = "http://localhost",
  port = 8080,
  catalog = "hive",
  schema = "default"
)
```

`dbConnect()` calls `/v1/info` before returning, so a wrong address or a
rejected credential fails immediately rather than on your first query. The
port can also be written into the host, as in `host = "localhost:8080"`.

## Authentication

Each authentication method is a closure that attaches credentials to every
request. The credentials live inside that closure, never in the global
environment, and never in a slot you might print by accident.

All three need an `https://` host. Over plain HTTP the credentials would
travel unencrypted, so `dbConnect()` refuses to send them there, before
anything reaches the network. Only if the connection is encrypted by other
means, such as an SSH tunnel to the coordinator, set `allow_http_auth = TRUE`.

### Basic — LDAP or a password file

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.example.com",
  port = 443,
  user = Sys.getenv("TRINO_USER"),
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_basic(
    Sys.getenv("TRINO_USER"),
    Sys.getenv("TRINO_PASSWORD")
  )
)
```

### JWT — service accounts and pipelines

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.example.com",
  port = 443,
  user = Sys.getenv("TRINO_USER"),
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_jwt(Sys.getenv("TRINO_TOKEN"))
)
```

### OAuth2 client credentials — corporate SSO

`httr2` caches the token and renews it when it expires, so a long-running
session does not need to think about the token's lifetime.

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.example.com",
  port = 443,
  user = Sys.getenv("TRINO_USER"),
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_oauth2(
    client_id     = Sys.getenv("OAUTH_CLIENT_ID"),
    client_secret = Sys.getenv("OAUTH_CLIENT_SECRET"),
    token_url     = "https://auth.example.com/oauth/token"
  )
)
```

## Internal certificate authorities

A cluster whose certificate is signed by your own CA does not need certificate
checking turned off. Point `trino_ssl()` at the CA's PEM file and verification
keeps working:

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.internal",
  port = 443,
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_basic(
    Sys.getenv("TRINO_USER"),
    Sys.getenv("TRINO_PASSWORD")
  ),
  ssl_options = trino_ssl(ca_bundle = "/etc/ssl/certs/internal-ca.pem")
)
```

`trino_ssl(verify = FALSE)` exists for a development cluster with a throwaway
certificate. It accepts *any* certificate, including one presented by an
attacker, so `dbConnect()` warns when it opens a connection with it.

## Querying

```{r}
sales <- dbGetQuery(con, "SELECT * FROM sales LIMIT 100")
```

For a result too large to hold in memory, take it in chunks. `dbFetch()`
follows Trino's pagination for you and buffers whatever a page delivered beyond
what you asked for:

```{r}
res <- dbSendQuery(con, "SELECT * FROM sales")
while (!dbHasCompleted(res)) {
  chunk <- dbFetch(res, n = 10000)
  # summarise, write out, accumulate...
}
dbClearResult(res)
```

Clearing a result that is still running sends a `DELETE` to the coordinator, so
abandoning a query also stops the cluster working on it. So does interrupting
R while it waits for a query.

`dbColumnInfo()` and `dbFetch(res, n = 0)` describe the result's columns, with
their types, before any row has been fetched.

## Statements

`dbExecute()` runs a statement to completion and returns how many rows Trino
says it changed: the count of an `INSERT`, `UPDATE`, `DELETE`, `MERGE` or
`CREATE TABLE ... AS SELECT`, and `0` for a statement that changes no rows,
such as `CREATE TABLE` or `DROP TABLE`.

```{r}
dbExecute(con, "INSERT INTO memory.default.visits VALUES (1, 'a'), (2, 'b')")
#> [1] 2
```

RTrino has no parameter binding, and passing `params` is an error rather than
being ignored. Put values into a statement with `sqlInterpolate()` or
`glue::glue_sql()`: both quote them through `dbQuoteLiteral()`, which writes
Trino literals of the right type, so a date is compared as a date and a string
cannot break out of its quotes.

```{r}
sql <- sqlInterpolate(
  con,
  "SELECT * FROM orders WHERE orderdate >= ?since AND orderstatus = ?status",
  since = as.Date("1995-01-01"),
  status = "F"
)
sql
#> <SQL> SELECT * FROM orders
#>       WHERE orderdate >= DATE '1995-01-01' AND orderstatus = 'F'
dbGetQuery(con, sql)
```

A `POSIXct` is written as a `TIMESTAMP` in the connection's
`session.timezone`, to the microsecond, which is how Trino reads it back.

RTrino reads data; it does not upload data frames. `dbWriteTable()`,
`dbAppendTable()`, `dbCreateTable()` and `dplyr::copy_to()` fail with a message
that says so: create and fill tables with SQL, for example
`CREATE TABLE ... AS SELECT ...`, through `dbExecute()`. Transactions
(`dbBegin()` and friends) are not supported either; each statement is committed
when it finishes.

## Timeouts

A query is a series of HTTP requests, and two limits apply:

* `timeout` (60 seconds by default) bounds each request, so a coordinator or a
  network that stops answering fails with an error instead of blocking R.
* `query_max_run_time` bounds the whole query on the cluster. Past it, Trino
  stops the query and RTrino raises a `trino_query_error` with
  `error_name = "EXCEEDED_TIME_LIMIT"`.

In a Shiny app, where a blocked R session blocks its user, set both:

```{r}
con <- dbConnect(
  Trino(),
  host = "https://trino.example.com",
  port = 443,
  catalog = "hive",
  schema = "analytics",
  auth = trino_auth_jwt(Sys.getenv("TRINO_TOKEN")),
  timeout = 30,
  query_max_run_time = "5m"
)
```

## Big integers and precision

Trino's `BIGINT` spans values an R double cannot hold exactly. By default
`RTrino` returns such columns as `bit64::integer64`, which is exact:

```{r}
dbGetQuery(con, "SELECT 9007199254740993 AS id")$id
#> integer64
#> [1] 9007199254740993
```

The one exception is the lowest `BIGINT`, -9223372036854775808, which bit64
keeps for its own `NA`: it comes back as `NA`, with a warning. Choose a
different representation at connection time with `bigint = "numeric"` (a
double, exact only up to 2^53) or `bigint = "character"` (the exact digits,
always).

`DOUBLE` values arrive bit for bit, `NaN` and the infinities included.
`DECIMAL` is returned as a double, so a decimal with more than 15 significant
digits is rounded; cast it to `VARCHAR` in the query to keep every digit.
`TIMESTAMP` and `TIME` keep their full precision, down to the microsecond.

## dplyr

With dbplyr 2.6.0 or newer installed, a connection carries a Trino SQL dialect
and `dplyr` verbs are translated to SQL Trino accepts:

```{r}
library(dplyr)

tbl(con, "sales") |>
  filter(year == 2026) |>
  group_by(region) |>
  summarise(total = sum(amount, na.rm = TRUE)) |>
  collect()
```

Trino-specific translations are applied so that an expression computes in
Trino what it computes in R:

* `/` divides as a double, where Trino would truncate the quotient of two
  integers, and `%%` takes the sign of the divisor, as in R.
* `as.integer()` and `as.integer64()` truncate towards zero, where Trino's cast
  would round.
* `paste()`, `paste0()` and `str_c()` accept columns of any type.
* `grepl()`, `sub()` and `gsub()` become `REGEXP_LIKE`/`REGEXP_REPLACE`,
  honour `fixed = TRUE` and `ignore.case = TRUE`, and accept R's `\\1`
  back-references; `sub()` replaces the first match only.
* A `Date` or `POSIXct` from R, as in `filter(orderdate >= !!since)`, becomes a
  typed Trino literal.

Where there is no faithful translation, a few differences remain, and it is
worth knowing them before moving a computation from R to the cluster:

* `round()` rounds halves away from zero, where R rounds them to the even
  digit: `round(2.5)` is `3` in Trino and `2` in R.
* `median()` and `quantile()` become `APPROX_PERCENTILE`, the percentile Trino
  offers over an arbitrary number of rows. It is approximate: on skewed data it
  can be off by a few per cent. Collect the rows and compute in R when the
  exact value matters.
* `paste()` skips `NA`, where R writes `"NA"`, and renders a `DOUBLE` column
  the way Trino does, so a value of 1.5 becomes `"1.5E0"`; round or cast it to
  a `DECIMAL` first when that matters.
* Regular expressions are Trino's (Java's syntax), not R's; the common
  constructs mean the same in both.
* `date + 1` fails, because Trino does not add a number to a date; write
  `date + days(1)`. `days()`, `weeks()`, `months()`, `years()`, `hours()`,
  `minutes()` and `seconds()` become intervals.

`compute()` needs `temporary = FALSE` and a name, because Trino has no
temporary tables.

`filter_out()`, added in dplyr 1.2.0, is translated with Trino's native
`IS DISTINCT FROM` instead of the portable `CASE WHEN` comparison dbplyr falls
back to, so the rows where the condition is `NA` are kept and the SQL stays
readable.

Two groups of dplyr 1.2.0 functions have no SQL translation in dbplyr 2.6.0 and
therefore none here either: `when_any()`/`when_all()`, which pass through
untranslated and reach the cluster as an unknown function, and
`recode_values()`/`replace_values()`/`replace_when()`, which fail with
"Cannot translate a `<formula>` object to SQL". Use `|` and `&` in place of
`when_any()` and `when_all()`, and `case_when()` or `if_else()` in place of the
recoding family, until dbplyr covers them.

Inspect the generated SQL before running it with `dplyr::show_query()`, and the
cluster's plan with `dplyr::explain()`. `simulate_trino()` gives a connection
that dispatches the Trino dialect without a cluster, for checking a translation
offline.

## Introspection

```{r}
dbListTables(con)
dbListFields(con, "sales")
dbExistsTable(con, "sales")
dbGetInfo(con)$db.version
```

A table can be named as `"table"`, `"schema.table"` or `"catalog.schema.table"`,
with `Id(catalog = , schema = , table = )`, or with an identifier that is
already quoted. Trino folds every identifier to lower case, so
`dbExistsTable(con, "SALES")` finds `sales`, as `SELECT * FROM SALES` would.

## Session properties and other headers

Anything the protocol expects in a header can be added to every request through
`extra.headers` — Trino session properties, for instance. `query_max_run_time`
joins whatever properties are set here:

```{r}
con <- dbConnect(
  Trino(),
  host = "http://localhost",
  port = 8080,
  catalog = "hive",
  schema = "default",
  extra.headers = list("X-Trino-Session" = "query_max_run_time=2h")
)
```

## Disconnecting

There is no server-side session to close, so `dbDisconnect()` only invalidates
the object; any later use of it is an error rather than a silent reconnect.

```{r}
dbDisconnect(con)
```
